Tracing the Meta-Level: PyPy's Tracing JIT Compiler

### 2009-07-04 talk/icooolps2009/talk/talk.tex
If we're using tex, we can go for a correct spelling
Small reorganizion of the slide "Idea for a Solution"

\title[PyPy's Tracing JIT Compiler]{
    Tracing the Meta-Level: PyPy's Tracing JIT Compiler
}
\author[Bolz, Cuni, Fijalkowski, Rigo]
{
    \textcolor{green!50!black}{Carl~Friedrich~Bolz}\inst{1} \and
    Antonio Cuni\inst{2} \and
    Maciej Fijalkowski\inst{3} \and
    Armin Rigo
}

@@

\begin{frame}
    \frametitle{Applying a Tracing JIT to an Interpreter}
    \begin{itemize}
    \item Question: What happens if the program is itself a bytecode interpreter?
    \item the (most important) hot loop of a bytecode interpreter is the bytecode dispatch loop
    \item Assumption violated: two iterations of the dispatch loop will usually take very different code paths
    \end{itemize}
    \pause
    \begin{block}{Terminology}
        \begin{itemize}
        \item \emph{tracing interpreter:} the interpreter that originally runs the program and produces traces
        \item \emph{language interpreter:} the bytecode interpreter runs on top
        \item \emph{user program:} the program run by the language interpreter
        \end{itemize}
    \end{block}
\end{frame}

@@

\begin{frame}
    \frametitle{Idea for a Solution}
    \begin{itemize}
    \item key idea: try to trace the loops in the user program
    \item approach: add things to the position key of the tracer
    \item tracing interpreter needs information about the language interpreter
    \item provided by adding \emph{hints} to the language interpreter
    \end{itemize}
    \pause
    \begin{block}{Hints Give Information About:}
    \begin{itemize}
    \item which variables make up the program counter of the language interpreter
    \item where the bytecode dispatch loop is
    \item which bytecodes can correspond to backward jumps
    \end{itemize}
    \end{block}
\end{frame}

@@

\begin{frame}
    \frametitle{What Have We Won?}
    \begin{itemize}
    \item trace corresponds to one loop of the user program
    \item however, most operations concerned with manipulating bytecode and program counter
    \item bytecode and program counter are part of the position key
    \item thus they are constant at the beginning of the loop
    \item therefore they can and should be constant-folded
    \end{itemize}
\end{frame}

@@

\begin{frame}
    \frametitle{Results}
    \begin{itemize}
    \item almost only computations related to the user program remain
    \item list of registers is only vestige of language interpreter
    \end{itemize}
    \pause
    \begin{block}{Timing Results Computing Square of 10'000'000}
    \begin{tabular}{|l|l|r|r|}
\hline
& &Time (ms) &speedup\\
\hline
1 &No JIT &442.7 $\pm$ 3.4 &1.00\\
2 &Normal Trace Compilation &1518.7 $\pm$ 7.2 &0.29\\
3 &Unrolling of Interp. Loop &737.6 $\pm$ 7.9 &0.60\\
4 &JIT, Full Optimizations &156.2 $\pm$ 3.8 &2.83\\
\hline
\end{tabular}
\end{block}
\end{frame}

>>>>

\title[PyPy's Tracing JIT Compiler]{
    Tracing the Meta-Level: PyPy's Tracing JIT Compiler
}
\author[Bolz, Cuni, Fijalkowski, Rigo]
{
    \textcolor{green!50!black}{Carl~Friedrich~Bolz}\inst{1} \and
    Antonio Cuni\inst{2} \and
    Maciej Fija\l{}kowski\inst{3} \and
    Armin Rigo
}

@@

\begin{frame}
    \frametitle{Applying a Tracing JIT to an Interpreter}
    \begin{itemize}
    \item Question: What happens if the program is itself a bytecode interpreter?
    \item the (most important) hot loop of a bytecode interpreter is the bytecode dispatch loop
    \item Assumption violated: consecutive iterations of the dispatch loop will usually take very different code paths
    \end{itemize}
    \pause
    \begin{block}{Terminology}
        \begin{itemize}
        \item \emph{tracing interpreter:} the interpreter that originally runs the program and produces traces
        \item \emph{language interpreter:} the bytecode interpreter run on top
        \item \emph{user program:} the program run by the language interpreter
        \end{itemize}
    \end{block}
\end{frame}

@@

\begin{frame}
    \frametitle{Idea for a Solution}
    \begin{itemize}
    \item goal: try to trace the loops \emph{in the user program,}
          and not just one iteration of the bytecode dispatch loop
    \item tracing interpreter needs information about the language interpreter
    \item provided by adding \emph{three hints} to the language interpreter
    \end{itemize}
    \pause
    \begin{block}{Hints Give Information About:}
    \begin{itemize}
    \item which variables make up the program counter of the language interpreter
    \item where the bytecode dispatch loop is
    \item which bytecodes can correspond to backward jumps
    \end{itemize}
    \end{block}
\end{frame}

@@

\begin{frame}
    \frametitle{What Have We Won?}
    \begin{itemize}
    \item trace corresponds to one loop of the user program
    \item however, most operations are concerned with manipulating bytecode and program counter
    \item bytecode and program counter are part of the position key
    \item thus they are constant at the beginning of the loop
    \item therefore they can and should be constant-folded
    \end{itemize}
\end{frame}

@@

\begin{frame}
    \frametitle{Results}
    \begin{itemize}
    \item almost only computations related to the user program remain
    \item list of registers is only vestige of language interpreter
    \end{itemize}
    \pause
    \begin{block}{Timing Results Computing Square of 10'000'000}
    \begin{tabular}{|l|l|r|r|}
\hline
& &Time (ms) &speedup\\
\hline
1 &No JIT &442.7 $\pm$ 3.4 &1.00\\
2 &JIT, Normal Trace Compilation &1518.7 $\pm$ 7.2 &0.29\\
3 &JIT, Unrolling of Interp. Loop &737.6 $\pm$ 7.9 &0.60\\
4 &JIT, Full Optimizations &156.2 $\pm$ 3.8 &2.83\\
\hline
\end{tabular}
\end{block}
\end{frame}
